A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-14 tex/codatadp.tex
Reorganize

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}

@@

  * Rewriter
  * Actual calculus

  * Correctness

@@

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a co-datatype or datatype type,
where a cyclic term $t$ describes an infeasible class of values in the case that $\tau \not\in \Codata$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is updated by applying the following rule exhaustively:

\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  C( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  C \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. C( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

@@

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$.
A possible sequence of applications of the above rule is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\hline
\end{array}\]
Informally, this indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( v_{\ec{w}}, \const{C}( v_{\ec{z}}, \ldots ))$ (resp. $\const{C}( v_{\ec{z}}, \const{C}( v_{\ec{w}}, \ldots ))$) 
for some value of $v_{\ec{w}}$ and $v_{\ec{z}}$.
$\Box$
\end{example}

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t_1}$ of datatype type whose class of values $\Val \ec{t_1} = \mu x. t_2$ are cyclic,
as indicated by the assumption $x \in FV( t_2 )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of co-datatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent modulo renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.

@@

\begin{comment}
To explain the rules in Figure~\ref{fig:ab-rules}, we introduce the following notions.
First, an edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is a transition between two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
(which we will call its \emph{source} and \emph{destination} respectively),
labeled by a constructor $C$ and integer $n$.
We say that edge $\ec{t_1} \rightarrow^{C}_n \ec{t_2}$ is \emph{induced by $\Ec$} if and only if $C( \vec u ) \in \ec{t_1}$ for some $\vec u$ and $\ec{t_2} = \ec{u_n}$.
In other words, $t_1$ is equivalent to a $C$-application whose $n^{th}$ argument is equivalent to $t_2$.
A \emph{path} $p$ is a non-empty ordered list of edges $e_1, \ldots e_n$ such that the destination of $e_i$ is the source of $e_{i+1}$ for $i = 1, \ldots (n-1)$,
and the sources of $e_1, \ldots, e_n$ are distinct.
We say that path $p$ is \emph{induced by $\Ec$} if and only if each of its edges are induced by $\Ec$.
Notice that since $\Ec$ contains a finite set of equivalence classes, all paths induced by $\Ec$ have finite length.
We call the \emph{source} of a path is the source of its first edge, and the \emph{destination} of a path is the destination of its last edge.
We say a path $p$ is \emph{maximal} if its destination $\ec{t}$ is either
the source of an edge in $p$, 
or does not contain an application of a non-nullary constructor (including the case when $t$ is not of datatype/codatatype type).
The set $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ is the smallest set containing all maximal paths with source $\ec{t_1}$ and destination $\ec{t_2}$ that are induced by $\Ec$.
The set $\tpath{\Ec}{\ec{t_1}}$ is the union of $\ttpath{\Ec}{\ec{t_1}}{\ec{t_2}}$ for all $t_2 \in \tEc$.
We will write, for instance, $\ec{t_1} \rightarrow^{C_1}_{n_1} \ec{t_2} \rightarrow^{C_2}_{n_2} \ec{t_3}$ 
as shorthand for the path containing an edge from $\ec{t_1}$ to $\ec{t_2}$, followed by one from $\ec{t_2}$ to $\ec{t_3}$.
%A set of paths is \emph{disjoint} if it does not contain two paths $p_1$ and $p_2$ such that $p_1 = p_2 \sigma$ for some $\sigma$.

The rule \rn{Cyclic} recognizes when a path exists from an equivalence class $\ec{t}$ to itself, as indicated
by the assumption $\ttpath{\Ec}{\ec{t}}{\ec{t}} \neq \emptyset$.
In this case, $\Ec$ entails that $t$ contains itself as a (strict) subterm,
and thus if $t$ is a datatype, $\Ec$ must be unsatisfiable.
The rule \rn{Bisimilar} recognizes when two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ having co-datatype type are observationally equivalent.
In this rule, $\sigma$ is a complete mapping over the equivalence classes that occur as sources in the edges of paths in $\tpath{\Ec}{\ec{t_2}}$.
The rule applies if $\tpath{\Ec}{\ec{t_1}}$ is comprised of the paths $p \sigma$ for each (and only) the paths $p \in \tpath{\Ec}{\ec{t_2}}$,
in which case $t_1$ must be equal to $t_2$.
We demonstrate an application of this rule concretely with the following example.
\end{comment}

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Singleton}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\\[3.7ex]
\rn{Split} 
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\end{tabular}
\caption{Rules for splitting.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Step 3: Branching}
If neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the degenerate case when there are two terms from $\tEc$ having a singleton codatatype type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

An application of a rule is \emph{redundant} if it is non-terminal and at least one branch in its conclusion does not add a new equality to $\Ec$.
A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes are obtained by 
a non-redundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no non-redundant application of a rule can be applied to it.
If there exists a closed derivation tree with root node $\Ec$, then we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, then we conclude that $\Ec$ is $\thD$-satisfiable.

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are observationally equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

>>>>

%\renewcommand{\vec}[1]{\mathbf{#1}}
\renewcommand{\vec}[1]{\bar #1}
\newcommand{\Ec}{\mathsf{E}}
\newcommand{\Fc}{\mathsf{F}}
\newcommand{\Gc}{\mathsf{G}}
\newcommand{\Ac}{\mathsf{A}}
\newcommand{\Dc}{\mathsf{D}}
\newcommand{\Rc}{\mathsf{R}}
\newcommand{\vrange}{\mathsf{range}}
\newcommand{\vdom}{\mathsf{dom}}
\newcommand{\tEc}{\mathcal{T}(\Ec)}
\newcommand{\tcEc}{\mathcal{T}^\ast(\Ec)}
\newcommand{\rn}[1]{\textsf{\small #1}}
\newcommand{\cvc}{\textsc{cvc}{\small 4}\xspace}
\newcommand{\teq}{\approx}
\newcommand{\tneq}{\not\teq}
\newcommand{\rem}[1]{\textcolor{red}{[#1]}}
%\newcommand{\is}[1]{is\text{-} #1}
\newcommand{\is}[1]{\const{is#1}}
\newcommand{\ror}{\quad \parallel \quad}
\newcommand{\tpath}[2]{\mathcal{P}_{ #2 \rightarrow \_ }( #1 )}
\newcommand{\ttpath}[3]{\mathcal{P}_{ #2 \rightarrow #3 }( #1 )}
\newcommand{\interp}[1]{[\![ #1 ]\!]}
\newcommand{\dpath}[3]{\delta^{#1}_{#2}( #3 )}
\newcommand{\ec}[1]{ [ #1 ] }
\newcommand{\M}{\mathcal{M}}
\newcommand{\Val}{\mathcal{A}}
\newcommand{\nf}[1]{{#1}\!\!\downarrow}
\newcommand{\vsim}{\sim}
\newcommand{\vsimv}[1]{\vsim_{#1}}
\newcommand{\tpos}[2]{#1\!\mid_{#2}}

@@

@@

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve a mapping $\Val$ from equivalence classes to terms containing $\mu$-bindings.
Recall that $\mu$-bindings can be used for representing (possibly cyclic) terms and values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $\const{c}$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some capture-avoiding renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
For instance, 
$\mu x. \const{C}( x ) \vsim \mu y. \const{C}( y )$ and
$\mu x, \const{D}( y, x ) \vsim \mu z. \const{D}( y, z )$, but
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{C}( x )$,
$\mu x. \const{D}( y, x ) \not\vsim \mu x. \const{D}( z, x )$ and
$\mu x. \const{D}( y, x ) \not\vsim \mu y. \const{D}( y, x )$.
Also, note that
two $\mu$-terms may represent the same value while being not bisimilar, as in
$\mu x. \const{C}( \mu y. \const{C}( x ) ) \not\vsim \mu x. \const{C}( x )$.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

Intuitively, a $\mu$-term $t$ of type $\tau$ describes a class of values of $\tau$, 
%either in the case when $\tau$ is a co-datatype or datatype type,
where a cyclic $t$ describes an infeasible class of values when $\tau \not\in \Codata$.
At the beginning of this step, 
we compute a mapping $\Val$ from equivalence classes to $\mu$-terms 
that describe the class of values that terms in that equivalence class can take in the models of $\Ec$. 
For each equivalence class $\ec{x}$, we associate a fresh variable $v_{\ec{x}}$ of the same type as $x$ not occurring in $\tEc$,
and initially say that $\Val$ contains $\ec{x} \mapsto v_{\ec{x}}$,
in other words, there are initially no constraints on the values for any equivalence class $\ec{x}$.
Then, $\Val$ is modified by applying the following update exhaustively:

\(
\inferrule{
  v_{\ec{x}} \in FV( \Val )
  \quad
  \const{C}( t_1, \ldots, t_n ) \in \ec{x}
  \quad
  \const{C} \in \Ctr
}{
  \Val := \Val \{ v_{\ec{x}} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{t_1}}, \ldots, v_{\ec{t_n}} ) \}
}
\)

@@

\begin{example}
Say that $\Ec$ contains (distinct) equivalence classes $\ec{x}$, $\ec{y}$, $\ec{z}$ and $\ec{w}$ 
where $\const{C}(z,x) \in \ec{y}$ and $\const{C}(w,y) \in \ec{x}$ for some $\const{C} \in \Ctr$.
A possible sequence of updates applied to $\Val$ is shown below.
In the first column, we show the equivalence class that was considered,
and in the second column, we show the content of $\Val$.
\[\begin{array}{|l|@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l|}
\hline
\text{Eqc} & \multicolumn{3}{c}{\Val} & \ 
\\
\hline
& \{ & \ec{x} \mapsto v_{\ec{x}}, & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto v_{\ec{y}}, & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{y}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\ec{x}
& \{ & \ec{x} \mapsto \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, v_{\ec{x}} ) ), & \ec{z} \mapsto v_{\ec{z}}, & \\
&    & \ec{y} \mapsto \mu v_{\ec{y}}. \const{C}( v_{\ec{z}}, \mu v_{\ec{x}}. \const{C}( v_{\ec{w}}, v_{\ec{y}} ) ), & \ec{w} \mapsto v_{\ec{w}} & \} \\
\hline
\end{array}\]
This indicates that the values for $x$ (resp. $y$) in models of $\Ec$
are of the form $\const{C}( v_{\ec{w}}, \const{C}( v_{\ec{z}}, \ldots ))$ (resp. $\const{C}( v_{\ec{z}}, \const{C}( v_{\ec{w}}, \ldots ))$) 
for some value of $v_{\ec{w}}$ and $v_{\ec{z}}$.
$\Box$
\end{example}

Given $\Val$, the rules for cyclicity and bisimilarity (shown in Figure~\ref{fig:ab-rules}) can be stated as follows.
For cyclicity, if there is an equivalence class $\ec{t_1}$ of datatype type whose class of values $\Val \ec{t_1} = \mu x. t_2$ are cyclic,
as indicated by the assumption $x \in FV( t_2 )$,
then we can conclude that $\Ec$ is unsatisfiable.
For bisimilarity, if there are two equivalence classes $\ec{t_1}$ and $\ec{t_2}$ 
of co-datatype type whose classes of values $\Val \ec{t_1}$ and $\Val \ec{t_2}$ are equivalent for a capture-avoiding renaming of $\mu$-bound variables,
then we can conclude $t_1 \teq t_2$.
We demonstrate the bisimilarity rule concretely with the following example.

@@

\begin{figure}[t]
\centering
\begin{tabular}{c}
\rn{Singleton}
\(
\inferrule{
  t, u : \tau \in \tEc
  \quad
  \tau \in \Codata,
  \mid \tau \mid = 1
}{
  \Ec := \Ec, t \teq u
}
\)
\\[3.7ex]
\rn{Split} 
\(
\inferrule{
  t : \tau \in \tEc 
  \quad 
  ( s( t ) \in \tEc, s \in \Sel^\tau ) 
  \text{ or } 
  ( \tau \in \Data, \tau \text{ finite } )
}{
  \Ec := \Ec, t \teq \const{C}_1( s^1_1( t ), \ldots, s^{a_1}_1( t ) ) \ror \ldots \ror \Ec := \Ec, t \teq \const{C}_n( s^1_n( t ), \ldots, s^{a_n}_n( t ) ) 
}
\)
\end{tabular}
\caption{Rules for branching.  
%All pairs of terms whose type has cardinality 1 are entailed to be equal (the degenerate case).
%Constructors must be assigned for all terms $t$ if has a selector is applied to it, or if $t$ has finite datatype type.
}
\label{fig:split-rule}
\end{figure}

\paragraph{Step 3: Branching.}
If neither a rule from Figure~\ref{fig:cc-rules} or Figure~\ref{fig:ab-rules} applies to $\Ec$, 
the rules from Figure~\ref{fig:split-rule} are applied.
The rule \rn{Singleton} handles the degenerate case when there are two terms from $\tEc$ having a singleton codatatype type,
in which case they must be equivalent.
Otherwise, in the case there is a selector term applied to $t$, or if $t$ has a finite datatype type, 
then the equivalence class of $t$ must contain an application of one of the constructors of $\tau$ (where $\Ctr^\tau = \{ C_1, \ldots C_n \}$), 
which is enforced by the rule \rn{Split}.

@@

An application of a rule is \emph{redundant} if it is non-terminal and at least one branch in its conclusion does not add a new equality to $\Ec$.
A \emph{derivation tree} is a tree whose nodes are sets of equalities, where non-root nodes vtained by 
a non-redundant application of a derivation rule to its parent node.
A derivation tree is \emph{closed} if all of its leaf nodes are $\bot$.
A node is \emph{saturated} if no non-redundant application of a rule can be applied to it.
If there exists a closed derivation tree with root node $\Ec$, then we conclude that $\Ec$ is $\thD$-unsatisfiable.
If there exists a derivation tree with root node $\Ec$ that contains a saturated node, then we conclude that $\Ec$ is $\thD$-satisfiable.

@@

\begin{lemma}[Refutation Soundness]%
\label{lem:rs}%
\afterDot
If there exists a closed derivation tree with root node $\Ec$, then $\Ec$ is unsatisfiable in $\thD$.
\end{lemma}
\begin{proof}
We prove this by structural induction on the (closed) derivation tree with root node $\Ec$.
First, if the derivation tree is an application of \rn{Conflict}, \rn{Unify$_2$}, or \rn{Cyclic},
then $\Ec$ is unsatisfiable in $\thD$.
For \rn{Conflict}, this is a consequence of equality reasoning.
For \rn{Unify$_2$}, this is a consequence of distinctness.
For \rn{Cyclic}, our construction of $\Val$ indicates that the class of values that $t_1$ can take in models of $\Ec$ is infeasible,
and thus $\Ec$ is unsatisfiable.
If the children of $\Ec$ are closed derivation trees 
whose root node is the result of applying the rule \rn{Split} on term $t$ of type $\tau$,
then by the induction hypothesis $\Ec \cup t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$ is unsatisfiable
for each $C_i \in \Ctr^\tau$.
Since by exhaustiveness, all models of $\thD$ entail (exactly) one equality $t \teq C_i( s^1_i( t ), \ldots, s^{a_i}_i( t ) )$,
we know $\Ec$ is unsatisfiable in $\thD$.
Otherwise, the child of $\Ec$ is a closed derivation tree 
whose root node is $\Ec \cup t_1 \teq t_2$ obtained by applying one of the rules \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, \rn{Unify$_1$}, \rn{Bisimilar}, or \rn{Singleton}.
In all cases, we have that $\Ec \models_{\thD} t_1 \teq t_2$.
For \rn{Refl}, \rn{Symm}, \rn{Trans}, \rn{Cong}, this is a consequence of equality reasoning.
For \rn{Unify$_1$}, this is a consequence of injectivity.
For \rn{Bisimilar}, our construction of $\Val$ indicates that the value of $t_1$ and $t_2$ are equivalent in all models of $\Ec$.
For \rn{Singleton}, clearly $t_1$ and $t_2$ must have the same value since the cardinality of their type is one.
By the induction hypothesis, we have $\Ec \cup t_1 \teq t_2$ is unsatisfiable in $\thD$, 
and thus $\Ec$ is unsatisfiable in $\thD$.
\qed
\end{proof}

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
We demonstrate a particular model $\M$ of $\thD$ that satisfies $\Ec$.
Before doing so, we require additional terminology concerning $\mu$-terms.

For a $\mu$-term $t_1$ with subterm $t_2$, 
the \emph{interpretation of $t_2$ in $t_1$} is the $\mu$-term $t_1\interp{t_2}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec u ) \in t_1$ indicates that $\mu x. C( \vec u )$ is the subterm of $t_1$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
t_1\interp{x}^M & = & \mu x. C( t_1\interp{ \vec u}^{M \cup \{ x \}}) & x \not\in M, x \mapsto \mu x. C( \vec u ) \in t_1  \\
t_1\interp{x}^M & = & x & \text{otherwise} \\
t_1\interp{\mu x. C( \vec u )}^M & = & \mu x. C( t_1\interp{ \vec u }^{M \cup \{ x \}} ) & x \not\in M \\
t_1\interp{\mu x. C( \vec u )}^M & = & x & \text{otherwise}
\end{array}\]
We will often write $t_1\interp{t_2}$ as shorthand for $t_1\interp{t_2}^\emptyset$.
We will write $\mathcal{T}( t_1 )$ to denote the set consisting of $t_1\interp{t_2}$ for all subterms $t_2$ of $t_1$,
and write $\mathcal{T}( S )$ to denote the union of $\mathcal{T}( t )$ for each $t \in S$ where $S$ is a set of terms.

Given a $\mu$-term $t_1$ of the form $\mu x. C( u_1, \ldots, u_n )$ with subterm $t_2$,
we say $t_2$ is a \emph{bisimilar subterm} of $t_1$ if and only if it is of the form
$\mu y. C( v_1, \ldots, v_n )$ and $t_1\interp{ u_i } \sim t_1\interp{ v_i }$ for $i = 1,\ldots, n$.
We say $t_1$ is \emph{normal} if and only if it does not contain a strict bisimilar subterm
and all of its strict subterms are also normal.
For instance, the $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( y ) )$ is not normal, 
since $\mu y. \const{S}( y )$ is a bisimilar subterm of $t_1$.
In particular, note that their arguments have the same interpretation in $t_1$, 
that is,
$t_1\interp{\mu y. \const{S}( y )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$ 
is syntactically equivalent to
$t_1\interp{y}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{y}^{\{y\}} ) =$ 
$\mu y. \const{S}( y )$.
The $\mu$-term $t_1$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is also not normal, 
since $\mu y. \const{S}( x )$ is a bisimilar subterm of $t_1$,
noting that
$t_1\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( t_1\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( t_1\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( t_1\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$
is syntactically equivalent to $t_1\interp{x}^\emptyset = t_1$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subterms of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.
